HiC Contact Map Comaprison Using Graphlet Approach

### 2018-03-27 papers/manuscripts/manuscript.tex
manuscript modification

We consider two measures of \textit{difference} when comparing contact
map graphlets across cell lines. 
The first measure is \textit{signature distance vectors} between
each contact map of two cell lines. 
For a pair of cells A and B, let 
$\mathbf{S}^A_{ij}$  and $\mathbf{S}^B_{ij}$ be their
signature matrices. The \textit{signature distance} of
contact map $\mathbf{C}_{i,j}$ between A and B is denoted
by $\mathbf{d}^{A,B}_{ij}$. $\mathbf{d}^{A,B}_{ij}$ 
is a vector of size $|V_{i,j}|$
and its elements $d^{A,B}_{i,j,l}$ are
calculated using the following formula from \cite{prvzulj2007biological}:

\begin{equation}
    d^{A,B}_{i,j,l} = 
    \frac{1}{73}\sqrt{\sum_{o=0}^{72}{t_{lo}^2}}
    \label{eq:distance_total}
\end{equation}

@@

The second measure of comparison that we use captures how 
similar two orbits are in terms of their count 
frequencies across loci between two contact maps. 
Each column in $S_{ij}$ can provide information
regarding the \textit{frequency distribution} of orbits throughout
the contact map $C_{ij}$. 
We can find how similar these distributions are to each other using
correlation measures.
These correlations are denoted by $\mathbf{m}^{A,B}_{i,j}$ and
can be calculate using
any plausible correlation measure. 
In this study, for each contact map,  we calculated
similarity between orbit distributions using Pearson's r 
correlation, which is computationally efficient.
However, pearson's r might not be able to capture
non-functional relationships between distributions. As a result, we
also used Maximal Information Coefficient (MIC) 
\cite{reshef2011detecting} in order to compare
correlations. MIC calculates mutual information (MI) between two
distributions, but utilizes dynamic programming in order adjust
bin sizes and numbers in order to achieve highest MI.
MIC values between two variables fall between 0 and 1,
with 0 meaning the two variables are completely independent
and 1 meaning one is dependant on the other.
We used both Pearson's r and MI in order to compare orbit
frequencies. Although results from both approaches were more
or less consistent, MIC showed higher robustness than Pearson's 
r method.

If MIC is used as correlation measure, each element of 
 $\mathbf{c}$ is calculated as below:
\begin{equation}
    m^{A,B}_{ijo} = MIC(\mathbf{S}^A_{ij.o}, \mathbf{S}^B_{ij.o})
    \label{eq:mic}
\end{equation}
Alternatively, if we use Pearson criterion we would have:
\begin{equation}
    m^{A,B}_{ijo} = Pearson(\mathbf{S}^A_{ij.o}, \mathbf{S}^B_{ij.o})
    \label{eq:pearson}
\end{equation}

@@

Result of pair-wise contact map graphlet distances can is
illustrated in Figure \ref{fig:orbit-distances}.
Each point on the graph is the
average of the graphlet distance vector 
($\bar{\mathbf{d}}^{A,B}_{i,j}$)
of the two cell lines specified in the legend.
 A one sided paired t-test
was conducted . The resulting p-values showed highly 
significant differeces from zero for all pairs of
cells. Detailed results of t-tests for each
contact map can be fount in supplementary materials.
For each contact map, each pair of cells are
order based on whether they are statistially
larger that the other or not. For example
for contact map $C_{i,j}$, the results 
of t-test is as follows:\\
\smallskip
{
    \small
zero  $<$  call4-mit  $<$     all-rl  $<$     mit-rl  $<$   call4-rl  $<$    all-mit  $=$  all-call4
}\\
\smallskip
which means that $\mathbf{d}^{\scriptscriptstyle CALL4, MIT}_{1,1}$ is statistically larger than 0,
but less than $\mathbf{d}^{\scriptscriptstyle ALL, RL}_{1,1}$ 
and so forth. We can also conclude that $\mathbf{d}^{\scriptscriptstyle ALL, CALL4}_{1,1}$ is not
statistically different from $\mathbf{d}^{\scriptscriptstyle ALL, MIT}_{1,1}$
between graphlets extracted from all contact maps. 
Refer to supplementary
material for results of all hypotheis tests.

@@

    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/orbit-distances_intra.png}
        \caption{}
        \label{fig:orbit-distances_intra}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/orbit-distances_intra_separate.png}
        \caption{}
        \label{fig:orbit-distances_intra_separate}
    \end{subfigure}
    \caption{   Pair-wise graphlet signature difference for
                (a) all 276 contact maps.
                (b) only the 23 intra-chromosomal contact maps.
                (c) all 276 contact maps as well as corresponding
                standard errors.
                Each point on a graph is the result of averaging all
                the distances acroos all loci of that contact map.
                As can be viewd, although inter-chromosomal graphlets 
                do not show significant differences, RL's inter-chromosomal
                graphlets seem to be closer to MIT's. Also in some contact
                maps ALL graphlets show greater distance from MIT graphlets
                than the other two cell lines.
             }
    \label{fig:orbit-distances}
\end{figure*}

We caclulated pair-wise MIC for each orbit in each of the 276
contact maps from MIT data and ALL, RL and CALL4 data separately. 
Figure \ref{fig:contact_maps_correlations}
shows average orbit correlations across all contact maps
($\bar{\mathbf{c}}^{\scriptscriptstyle A,B}_{i,j}$: average along the red \textit{vertical} arrow in figure 
\ref{graphlet_correlation_schema}), while
figure \ref{fig:orbits_correlations} 
demonstrates average correlations across all 72 orbits within each  contact map
(average along the red \textit{horizontal} arrows in figure 
\ref{graphlet_correlation_schema}).
It is worth mentioning that \textit{interchromosomal} thresholded contact maps 
represent
a biparticde graph with the loci from each chromosome on one side. Due to this
bipartide nature of the graphs in inter-chromosomal maps,
count of certain orbits is always 0, resulting in
a correlation values of 0 for them as well.
We ignored these values  when we calculated averages across orbits 
in figure \ref{fig:contact_maps_correlations_all} since they
would result in a bias towards zero in averages. You can see the bias in 
figure \ref{fig:orbits_correlations_all} where average correlations of obits
3, 9, 10-14, 20-34, 39-48, and 51-72 are close to zero. In fact all correlations
corresponding to these orbits are 0 except for the ones between the same 
chromosomes which are illustrated in Figure 
\ref{fig:orbits_correlations_intra}.

\begin{figure*}[t]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/contact_maps_correlations_all.png}
        \caption{}
        \label{fig:contact_maps_correlations_all}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/contact_maps_correlations_intra.png}
        \caption{}
        \label{fig:contact_maps_correlations_intra}
    \end{subfigure}
    \caption{pair-wise average contact map orbit correlations
             (a) for all contact maps
             (b) only for intra-chromosomal contact maps
             These values are calculated by averaging over 
             pairwise correlations of all orbits in a 
             contact map.
             }
    \label{fig:contact_maps_correlations}
\end{figure*}

\begin{figure*}[t]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth, height=.2\paperheight]{figures/orbits_correlations_all.png}
        \caption{}
        \label{fig:orbits_correlations_all}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth, height=.2\paperheight]{figures/orbits_correlations_intra.png}
        \caption{}
        \label{fig:orbits_correlations_intra}
    \end{subfigure}
    \caption{pair-wise average orbit correlations.
             In figure \ref{fig:orbits_correlations_all}, each point
             in the graph is the result of averaging pair-wise
             orbit correlations over all contact maps, while
             each point in figure \ref{fig:orbits_correlations_intra}
             are averaged only over intra-chromosomal contact maps.
             Counts for certain orbits are always zero in inter-chromosomal
             maps, leading to average value close to zero in 
             \ref{fig:orbits_correlations_all}.
             }
    \label{fig:orbits_correlations}
\end{figure*}
We have conducted one-sided
t-test in order to test whether the average correlations across
contact maps is equal to 1 (Equation \ref{eq:h2a}) and whether the average 
correlations across orbits is equal to 1 (Equation \ref{eq:h2b}). The results
for both test showed that all values are significantly less than 1.
Please refer to supplementary material for result of the t-tests.
\subsection{Comparing Cancer cells}
\begin{figure*}[t]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/orbit-distances_intra_cancer.png}
        \caption{}
        \label{fig:orbit_distances_intra_cancer}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/orbits_correlations_intra_cancer.png}
        \caption{}
        \label{fig:orbits_correlations_intra_cancer}
    \end{subfigure}
    \caption{}
    \label{fig:orbits_cancer_intra}
\end{figure*}

We also conducted the same comparison with the same hypothesis among cancer cells.
The results for intra-chromosomal distances and correlations can be found in
figure \ref{fig:orbits_cancer_intra}. As can be noticed, CALL4 and RL cell
line exhibit a significantly higher average correlations for the majority of
the chromosomes (figure \ref{fig:orbit_distances_intra_cancer}) and almost in
all orbits (figure \ref{fig:orbits_correlations_intra_cancer}).
\section{discussion}
\newpage
\section{Resources}
\textbf{Hi-C Datasets:}
\begin{enumerate}
    \item \href{https://github.com/rasoolianbehnam/watson}{Code base for this article}
    \item \href{http://sysbio.rnet.missouri.edu/T0510/tmp_download/link_to_download_genome_data/}
        {Datasets including cancerous cells}
    \item \href{https://bcm.app.box.com/v/aidenlab/folder/11234760671}{Original Datasets}
\end{enumerate}

>>>>

We consider two measures of \textit{difference} when comparing contact
map graphlets across cell lines. 
The first measure is \textit{signature distance vectors} between
each contact map of two cell lines. 
For a pair of cells A and B, let 
$\mathbf{S}^A_{ij}$  and $\mathbf{S}^B_{ij}$ be their
signature matrices. The \textit{signature distance} of
contact map $\mathbf{C}_{i,j}$ between A and B is denoted
by $\mathbf{d}^{\scriptscriptstyle A,B}_{ij}$. $\mathbf{d}^{\scriptscriptstyle A,B}_{ij}$ 
is a vector of size $|V_{i,j}|$
and its elements $d^{\scriptscriptstyle A,B}_{i,j,l}$ are
calculated using the following formula from \cite{prvzulj2007biological}:

\begin{equation}
    d^{\scriptscriptstyle A,B}_{i,j,l} = 
    \frac{1}{73}\sqrt{\sum_{o=0}^{72}{t_{lo}^2}}
    \label{eq:distance_total}
\end{equation}

@@

The second measure of comparison that we use captures how 
similar two orbits are in terms of their count 
frequencies across loci between two contact maps. 
Each column in $S_{ij}$ can provide information
regarding the \textit{frequency distribution} of orbits throughout
the contact map $C_{ij}$. 
We can find how similar these distributions are to each other using
correlation measures.
These correlations are denoted by $\mathbf{m}^{\scriptscriptstyle A,B}_{i,j}$ and
can be calculate using
any plausible correlation measure. 
In this study, for each contact map,  we calculated
similarity between orbit distributions using Pearson's r 
correlation, which is computationally efficient.
However, pearson's r might not be able to capture
non-functional relationships between distributions. As a result, we
also used Maximal Information Coefficient (MIC) 
\cite{reshef2011detecting} in order to compare
correlations. MIC calculates mutual information (MI) between two
distributions, but utilizes dynamic programming in order adjust
bin sizes and numbers in order to achieve highest MI.
MIC values between two variables fall between 0 and 1,
with 0 meaning the two variables are completely independent
and 1 meaning one is dependant on the other.
We used both Pearson's r and MI in order to compare orbit
frequencies. Although results from both approaches were more
or less consistent, MIC showed higher robustness than Pearson's 
r method.

If MIC is used as correlation measure, each element of 
 $\mathbf{c}$ is calculated as below:
\begin{equation}
    m^{\scriptscriptstyle A,B}_{ijo} = MIC(\mathbf{S}^A_{ij.o}, \mathbf{S}^B_{ij.o})
    \label{eq:mic}
\end{equation}
Alternatively, if we use Pearson criterion we would have:
\begin{equation}
    m^{\scriptscriptstyle A,B}_{ijo} = Pearson(\mathbf{S}^A_{ij.o}, \mathbf{S}^B_{ij.o})
    \label{eq:pearson}
\end{equation}

@@

Result of pair-wise contact map graphlet distances can is
illustrated in Figure \ref{fig:orbit-distances}.
Each point on the graph is the
average of the graphlet distance vector 
($\bar{\mathbf{d}}^{\scriptscriptstyle A,B}_{i,j}$)
of the two cell lines specified in the legend.
 A one sided paired t-test
was conducted . The resulting p-values showed highly 
significant differeces from zero for all pairs of
cells. Detailed results of t-tests for each
contact map can be fount in supplementary materials.
For each contact map, each pair of cells are
order based on whether they are statistially
larger that the other or not. For example
for contact map $C_{i,j}$, the results 
of t-test is as follows:
\vspace{.2cm}\\
\small{
zero  $<$  call4-mit  $<$     all-rl  $<$     mit-rl  $<$   call4-rl  $<$    all-mit  $=$  all-call4
}
\vspace{.2cm}\\
which means that $\mathbf{d}^{\scriptscriptstyle CALL4, MIT}_{1,1}$ is statistically larger than 0,
but less than $\mathbf{d}^{\scriptscriptstyle ALL, RL}_{1,1}$ 
and so forth. We can also conclude that $\mathbf{d}^{\scriptscriptstyle ALL, CALL4}_{1,1}$ is not
statistically different from $\mathbf{d}^{\scriptscriptstyle ALL, MIT}_{1,1}$
between graphlets extracted from all contact maps. 
Refer to supplementary
material for results of all hypotheis tests.

@@

    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/orbit-distances_intra.png}
        \caption{}
        \label{fig:orbit-distances_intra}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/orbit-distances_intra_separate.png}
        \caption{}
        \label{fig:orbit-distances_intra_separate}
    \end{subfigure}
    \caption{   Pair-wise graphlet signature difference for
                (\textbf{a}) all 276 contact maps
                ($\bar{\mathbf{d}}^{\scriptscriptstyle A,B}_{i,j} \quad
                \forall i,j \in \{1 ... 23\} \quad \& \quad j \ge i$).
                (\textbf{b}) only the 23 intra-chromosomal contact maps
                ($\bar{\mathbf{d}}^{\scriptscriptstyle A,B}_{i,i}$).
                (\textbf{c}) all 276 contact maps as well as corresponding
                standard errors.
                Each point on a graph is the result of averaging all
                the distances acroos all loci of that contact map.
                As can be viewd, although inter-chromosomal graphlets 
                do not show significant differences, RL's inter-chromosomal
                graphlets seem to be closer to MIT's. Also in some contact
                maps ALL graphlets show greater distance from MIT graphlets
                than the other two cell lines.
             }
    \label{fig:orbit-distances}
\end{figure*}

We caclulated pair-wise MIC for each orbit in each of the 276
contact maps from MIT data and ALL, RL and CALL4 data separately. 
Figure \ref{fig:contact_maps_correlations}
shows average orbit correlations across all contact maps, while
figure \ref{fig:orbits_correlations} 
demonstrates average correlations across all 72 orbits within each  contact map.
It is worth mentioning that \textit{interchromosomal} thresholded contact maps 
represent
a bipartide graph with the loci from each chromosome on one side. Due to this
bipartide nature of the graphs in inter-chromosomal maps,
count of certain orbits is always 0, resulting in
a correlation values of 0 for them as well.
We ignored these values  when we calculated averages across orbits 
in figure \ref{fig:contact_maps_correlations_all} since they
would result in a bias towards zero in averages. You can see the bias in 
figure \ref{fig:orbits_correlations_all} where average correlations of orbits
$\mathbb{Q} = \{3, 9, 10{\text -}14, 20{\text -}34, 39{\text -}48, 51{\text -}72\}$ 
are close to zero. In fact all correlations
corresponding to these orbits are 0 except for the ones between the same 
chromosomes which are illustrated in Figure 
\ref{fig:orbits_correlations_intra}.

\begin{figure*}[t]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/contact_maps_correlations_all.png}
        \caption{}
        \label{fig:contact_maps_correlations_all}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/contact_maps_correlations_intra.png}
        \caption{}
        \label{fig:contact_maps_correlations_intra}
    \end{subfigure}
    \caption{pair-wise average contact map orbit correlations
             (a) for all contact maps($\bar{\mathbf{m}}^{\scriptscriptstyle \scriptscriptstyle A,B}_{i,j} 
            \quad \forall i,j \in \{1 ... 23\} \quad \& \quad j \ge i$,
            average along the red \textit{vertical} arrow in figure 
            \ref{graphlet_correlation_schema})
             (b) only for intra-chromosomal contact maps
            ($\bar{\mathbf{m}}^{\scriptscriptstyle \scriptscriptstyle A,B}_{i,i} 
            \quad \forall i \in \{1 ... 23\} $)
             These values are calculated by averaging over 
             pairwise correlations of all orbits in a 
             contact map.
             }
    \label{fig:contact_maps_correlations}
\end{figure*}

\begin{figure*}[t]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth, height=.2\paperheight]{figures/orbits_correlations_all.png}
        \caption{}
        \label{fig:orbits_correlations_all}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth, height=.2\paperheight]{figures/orbits_correlations_intra.png}
        \caption{}
        \label{fig:orbits_correlations_intra}
    \end{subfigure}
    \caption{pair-wise average orbit correlations.
             In figure \ref{fig:orbits_correlations_all}, each point
             in the graph is the result of averaging pair-wise
             orbit correlations over all contact maps
             ($\frac{1}{276}\sum_{i=0}^{23}\sum_{j=i}^{23}{
             m^{\scriptscriptstyle \scriptscriptstyle A,B}_{i,j, o}} \quad 
             \forall o \in \{0, 1, ..., 72\}$,
             average along the red \textit{horizontal} arrows in figure 
             \ref{graphlet_correlation_schema}), while
             each point in figure \ref{fig:orbits_correlations_intra}
             are averaged only over intra-chromosomal contact maps
             ($\frac{1}{23}\sum_{i=0}^{23}{
             m^{\scriptscriptstyle \scriptscriptstyle A,B}_{i,i, o}} \quad 
         \forall o \in \{0, 1, ..., 72\}$).
             Counts for certain orbits are always zero in inter-chromosomal
             maps, leading to average value close to zero in 
             \ref{fig:orbits_correlations_all}.
             }
    \label{fig:orbits_correlations}
\end{figure*}
We have conducted one-sided
t-test in order to test whether the average correlations across
contact maps is equal to 1  and whether the average 
correlations across orbits is equal to 1. The results
for both test showed that all values are significantly less than 1.
Please refer to supplementary material for result of the t-tests.

\section{discussion}
\newpage
\section{Resources}
\textbf{Hi-C Datasets:}
\begin{enumerate}
    \item \href{https://github.com/rasoolianbehnam/watson}{Code base for this article}
    \item \href{http://sysbio.rnet.missouri.edu/T0510/tmp_download/link_to_download_genome_data/}
        {Datasets including cancerous cells}
    \item \href{https://bcm.app.box.com/v/aidenlab/folder/11234760671}{Original Datasets}
\end{enumerate}
